$\pi$ with leftovers: \\ a mechanisation in Agda

### 2021-01-06 paper/paper.tex
Shorten example in syntax

  \item We \textbf{generalise resource counting to a set of usage algebras that can be mixed within the same type system}.
    We do not instantiate our type system to only work with linear resources, instead we present an algebra-agnostic type system, and admit a mix of user-defined \emph{resource aware} algebras (\autoref{multiplicities}).
    Any \emph{partial commutative monoid} that is \emph{decidable}, \emph{deterministic}, \emph{cancellative} and has a \emph{minimal element} is a valid such algebra.
    Multiple algebras can be mixed in the type system --- usage contexts keep information about what algebra to for each type (\autoref{contexts}).
    In particular, this allows for type systems combining linear (use exactly once), graded (exact number of times $n$) and shared (free to reuse and discard) types under the same framework.
\end{itemize}

@@

% \todo{introduce shared and graded types sooner}
%   \item We use \textbf{type-level {de Bruijn indices} \cite{deBruijn1972,Dybjer1994} to introduce a syntax for \picalc{} processes that is well-scoped by construction}:
%     Every free variable is accounted for in the type of the process that uses it (\autoref{syntax}).
%     \todo{Do Ciccone \& Padovani also use type level de Bruijn indices?}
% 
%   \item We \textbf{parametrise our type system over a set of resource-aware usage algebras}:
%     A \emph{usage algebra} is defined as a ternary relation that is \emph{partial}, \emph{deterministic}, \emph{cancellative}, \emph{associative} and \emph{commutative}.
%     The user is able to choose any mix of algebras for the type system --- as long as they comply with the specification.
%     This allows for linear, graded and shared types to be mixed in the same type system.
% 
%   \item We \textbf{apply leftover typing to the \picalc{} and prove type preservation for the resulting type system}:
%     \todo{along the lines of Allais \cite{Allais2018a}}.
% \end{enumerate}
% \item \textbf{Formalisation of the syntax and the semantics of the \picalc{}}:
%  \begin{itemize}
%    \item \textbf{Syntax}: we use type level {de Bruijn indices} \cite{deBruijn1972,Dybjer1994} to introduce a syntax of \picalc{} processes that is \emph{well scoped by construction}: every free variable is accounted for in the type of the process that uses it (\autoref{syntax}).
%    For convenience and readability, the user can convert from and to a syntax using channel names (\autoref{from_to_deBruijn}).
%    These conversion functions are shown to be inverses of one another up to $\alpha$-conversion.
%    
%    \item \textbf{Semantics}: we provide an operational semantics for the \picalc{}, prior to any typing (\autoref{semantics}).
%    Thanks to our well-scoped grammar  (\autoref{syntax}), the operational semantics is defined on \emph{the totality of the syntax} as a reduction relation on processes (\autoref{operational-semantics}).
%    The reduction relation tracks at the type level the channel on which communication occurs. This information is later used to state the subject reduction theorem --- aka type preservation (\autoref{thm:subject-reduction}).
%    The reduction relation is defined modulo \emph{structural congruence} (\autoref{structural-congruence}) --- a relation defined on processes that acts as a quotient type to remove unnecessary syntactic minutiae introduced by the syntax of the \picalc{}.
%  \end{itemize}
%   
%   \item \textbf{Leftover typing for the \picalc{} with support for linear, graded and shared types}:
%   we present our type system for a resource-aware \picalc{} in \autoref{type-system}.
%   \begin{itemize}
%     \item \textbf{Algebras on multiplicities}: The user can provide \emph{resource-aware} algebras, which are then applied to the type system (\autoref{multiplicities}).
%     Any \emph{partial commutative monoid} that is \emph{decidable}, \emph{deterministic}, \emph{cancellative} and has a \emph{minimal element} is a valid such algebra --- linear types, graded types and shared types all satisfy these algebraic laws.
% 
%     Multiple algebras can be simultaneously used in a single type system --- usage contexts keep information about what algebra to use on which element (\autoref{contexts}).
%     This allows for type systems combining linear, graded and shared types under the same framework.
%     
%     \item \textbf{Leftover typing}: Our type system uses \emph{leftover typing} to model the resource-aware \picalc{} (\autoref{leftover-typing}).
%     This approach adds a leftover usage context to the typing judgements.
%     Typing derivations take the resources of their input usage context, consume some of them, and leave the remaining as leftovers in the output usage context.
%     Leftover typing \textbf{makes top-down context splits unnecessary}, allows for \emph{framing} (\autoref{thm:framing}) to be stated, and makes it possible to generalise \emph{weakening} (\autoref{thm:weakening}) and \emph{strengthening} (\autoref{thm:strengthening}) for the \picalc{}.
%   \end{itemize}
% 
%   \item \textbf{Fully mechanised formalisation and type safety results}:
%     The formalisation of the \picalc{} with leftover typing, from the syntax to the semantics and the type system, has been fully mechanised in Agda and is publicly available in our GitHub repository \cite{Zalakain2020Agda}
% 
%     We present the type safety results of our \picalc{} with leftovers in \autoref{type-safety}.
%     We have fully mechanised \emph{framing} (\autoref{thm:framing}), \emph{weakening} (\autoref{thm:weakening}) and \emph{strengthening} (\autoref{thm:strengthening}).
%     We use these results to then prove and mechanise \emph{subject congruence} (\autoref{thm:subject-congruence}), and together with \emph{substitution} (\autoref{thm:substitution}), to prove and mechanise \emph{subject reduction} (\autoref{thm:subject-reduction}).
% \end{enumerate}

@@

A variable reference occurring under $n$ binders can refer to $n$ distinct variables.
References outside of that range are meaningless.
It is useful to rule out these ill-scoped terms syntactically.
We do so by introducing the indexed family of types \cite{Dybjer1994} $\Var_n$: for all naturals $n$, the type $\Var_n$ has $n$ distinct elements.
We index processes according to their \emph{depth}: for all naturals $n$, a process of type $\Process_n$ contains variables that can refer to $n$ distinct elements.
Every time we go under a binder, we increase the index of the continuation process, allowing the variable references within to refer to one more thing.
\begin{mathpar}
  \datatype
  {n : \N}
  {\Var_n : \Set}
  \; \textsc{Var}

@@

\todo{adapt to de bruijn}
Process $\PO$ denotes the terminated process, where no further communications can occur.
Process $\new{}\; P$ creates a new channel bound with scope $P$.
Process $\comp{P}{Q}$ is the parallel composition of processes $P$ and $Q$.
Processes $\recv{x}{} P$ and $\send{x}{y} P$ denote respectively, the input and output processes of a variable $y$ over a channel $x$, with continuation $P$.
Scope restriction $(\new{}x) \; P$ and input $\recv{x}{y} \; P$ are \emph{binders}, they are the only constructs that introduce bound names --- $x$ and $y$ in $P$, respectively.

@@

We present an example that highlights the ways in which linear, graded and shared types can be used to prevent errors in a concurrent and distributed system defined by \picalc{} processes in parallel.

A courier system consists of three \emph{roles}:
a \sender{}, who wants to send a package;
a \receiver{}, who receives the package sent by the sender;
and a \courier{}, who carries the package from the sender to the receiver.

Our courier system (given in \autoref{fig:example-process}) is defined by four \picalc{} processes in parallel instantiating the above three roles:
we have two \sender{} processes, $S_1$ and $S_2$, sending data over channels $x$ and $y$, respectively;
one \receiver{} process, $R$, which receives over channel $z$ the data sent from each of the senders -- hence receives twice;
and a \courier{} process $C$, which synchronises communication among $S_1$, $S_2$ and $R$.
The \courier{} process $C$ first receives data from $S_1$ and $S_2$ on its input channels $x$ and $y$, respectively, and then sends the received data to $R$ along its output channel $z$.

We will now define the courier system as \picalc{} processes.
In our work, the syntax of \picalc{} processes adopts \emph{de Bruijn indices} \cite{deBruijn1972} for channel names and variables (details in \autoref{syntax}).
In short, for the purposes of this example:
process $\new{} P$ creates a new channel and makes it available at index $0$ in the continuation process $P$;
$\comp{P}{Q}$ composes processes $P$ and $Q$ in parallel; process $\recv{x}{}P$ receives data along channel $x$ and makes that data available at index $0$ in the continuation process $P$;
process $\send{x}{y}{P}$ sends variable $y$ over channel $x$ and continues as process $P$.
Both variable references $\Var_n$ and processes $\Process_n$ are indexed by the number $n$ of free variables they can refer to.
\begin{center}
  \small
  \begin{tikzpicture}[draw,circle]
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=brown] (send-x) at (0,2)   {$\func{send} \, x$};
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=brown] (send-y) at (0,0)   {$\func{send} \, y$};
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=purple] (recv-z) at (4,1)   {$\func{recv} \, z$};
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=cyan,align=center] (sync) at (2,1)   {$\func{carry}$\\$ \, x \, y \, z$};

@@

For the sake of clarity, we will define the four subprocesses in our courier system separately and parametrise them by the channels on which they operate.
Each role \sender{}, \receiver{} and \courier{} is defined as a function respectively, $\func{send}, \func{recv}$ and $\func{carry}$.

The \sender{} \emph{role} is defined below.
The \sender{} creates a new channel to be sent as data, and sends it over channel $c$, and then terminates, denoted by process $\PO$.
\begin{flalign*}
  & \func{send} \; : \; \Var_n \to \Process_n & \\
  & \func{send} \; c = \new{} \send{(\suc c)}{\constr{0}} \PO
\end{flalign*}
Each \sender{} \emph{process} $S_1$ and $S_2$ is an instantiation of $\func{send} \; c$.
In particular,
$S_1$ is defined as $\func{send} \; x$ and $S_2$ as $\func{send} \; y$, where data is sent over channels $x$ and $y$, respectively (see \autoref{fig:example-process}).

The \receiver{} role is defined below, where $\func{recv} \; c \;$ receives data \emph{twice} on a channel $c$ and then terminates with $\PO$.
\begin{flalign*}
  & \func{recv} \; : \; \Var_n \to \Process_n & \\
  & \func{recv} \; c \; = \recv{c}{} \recv{(\suc c)} \PO
\end{flalign*}
The \receiver{} process $R$ is defined as  $\func{recv} \; z$, with channel $z$ instantiating $c$, which is the channel used to receive data from \courier{} (see \autoref{fig:example-process}).

The \courier{} role is defined below.
The \courier{} process $C$ is defined as $\func{carry} \; x \; y \; z$.
It sequentially receives on the two input channels $x$ and $y$ -- instantiating $in0$ and $in1$, and then outputs the two pieces of received data on the output channel $z$ -- instantiating $out$ (see \autoref{fig:example-process}).
$$
\func{carry} \; : \; \Var_n \to \Var_n \to \Var_n \to \Process_n
$$
\begin{alignat}{5}
  \func{carry} \; in0 \; in1 \; out \; &&= \\
  \recv{in0 &&}{} \\
  \recv{(\suc \; in1) &&}{} \\
  \send{(\suc \suc \; out) &&}{\suc \constr{0}} \\
  \send{(\suc \suc \; out) &&}{\constr{0}} \PO
\end{alignat}

Finally, we compose these processes together and create three communication channels: the first channel is shared between the \sender{} $S_1$ and the \courier{} $C$, the second between the \sender{} $S_2$ and the \courier{} $C$, and the third between the \receiver{} $R$ and the \courier{} $C$.
The result is the courier systems defined below.
\begin{flalign*}
  & \func{system} && : \; \Process_{\constr{0}} & \\
  & \func{system} && = \new{} ( \func{send} \; \constr{0} \\
  & && \comp{}{ \new{} ( \func{send} \; \constr{0} } \\
  & && \comp{}{ \new{} ( \func{recv} \; \constr{0} } \\
  & && \comp{}{ \func{carry} \; (\suc \suc \constr{0}) \; (\suc \constr{0}) \; \constr{0}))) } \\
\end{flalign*}

In order to type the courier system we will use all three typing disciplines: linear, graded and shared typing.
We will continue this example and provide the typing derivations in \autoref{leftover-typing}.
We use \textbf{linear types} for the \emph{usage annotations} of data and \textbf{shared types} for the \emph{data} itself that is being sent from the \sender{} processes to the \receiver{}.
We use \textbf{graded types} for the usage annotations of the communication channels $x$, $y$, and $z$: the \sender{} processes $\func{send} \; x$ and $\func{send} \; y$ each consume $1$ output multiplicity and the \receiver{} process $\func{recv} \; z$ consumes $2$ input multiplicities, as it receives twice on channel $z$.
Dually, the \courier{} process $\func{carry} \; x \; y \; z$ consumes one input multiplicity on $x$, one input multiplicity on $y$, and two output multiplicities on $z$.
\end{example}

>>>>

  \item We \textbf{generalise resource counting to a set of usage algebras that can be mixed within the same type system}.
    We do not instantiate our type system to only work with linear resources, instead we present an algebra-agnostic type system, and admit a mix of user-defined \emph{resource aware} algebras (\autoref{multiplicities}).
    Any \emph{partial commutative monoid} that is \emph{decidable}, \emph{deterministic}, \emph{cancellative} and has a \emph{minimal element} is a valid such algebra.
    Multiple algebras can be mixed in the type system --- usage contexts keep information about what algebra to for each type (\autoref{contexts}).
    In particular, this allows for type systems combining linear (use exactly once), graded (exact number of times $n$) and shared (free to reuse and discard) types under the same framework.
\end{itemize}
\todo{mention we don't support recursion, why}

@@

@@

A variable reference occurring under $n$ binders can refer to $n$ distinct variables.
References outside of that range are meaningless, and it is useful to rule them out syntactically.
We do so by introducing the indexed family of types \cite{Dybjer1994} $\Var_n$: for all naturals $n$, the type $\Var_n$ has $n$ distinct elements.
We index processes according to their \emph{depth}: for all naturals $n$, a process of type $\Process_n$ contains free variables that can refer to $n$ distinct elements.
Every time we go under a binder, we increase the index of the continuation process, allowing the variable references within to refer to one more thing.
\begin{mathpar}
  \datatype
  {n : \N}
  {\Var_n : \Set}
  \; \textsc{Var}

@@

Process $\PO$ denotes the terminated process, where no further communications can occur;
process $\new{} P$ creates a new channel and binds it at index $0$ in the continuation process $P$;
$\comp{P}{Q}$ composes processes $P$ and $Q$ in parallel;
process $\recv{x}{}P$ receives data along channel $x$ and makes that data available at index $0$ in the continuation process $P$;
process $\send{x}{y}{P}$ sends variable $y$ over channel $x$ and continues as process $P$.

@@

As an example use of this syntax, we present a courier system that consists of three \emph{roles}:
a \sender{}, who wants to send a package;
a \receiver{}, who receives the package sent by the sender;
and a \courier{}, who carries the package from the sender to the receiver.
\begin{center}
  \small
  \begin{tikzpicture}[draw,circle]
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=brown] (send-x) at (0,2)   {$\func{send} \, x$};
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=brown] (send-y) at (0,0)   {$\func{send} \, y$};
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=purple] (recv-z) at (4,1)   {$\func{recv} \, z$};
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=cyan,align=center] (sync) at (2,1)   {$\func{carry}$\\$ \, x \, y \, z$};

@@

\todo{the colors are confusing: function color for send recv carry, individual colors for sender receiver and carrier}
Our courier system is defined by four \picalc{} processes composed in parallel instantiating the above three roles:
we have two \sender{} processes, $\func{send} \, x$ and $\func{send} \, y$, sending data over channels $x$ and $y$, respectively;
one \receiver{} process, $\func{recv} \, z$, which receives over channel $z$ the data sent from each of the senders -- hence receives twice;
and a \courier{} process $\func{carry} \, x \, y \, z$, which synchronises communication among the senders and the receiver.
The \courier{} process first receives data from the two senders along its input channels $x$ and $y$, and then sends the two received bits of data to the receiver along its output channel $z$.

The \sender{} and \receiver{} \emph{roles} are defined below, parametrised by the channels on which they operate.
The \sender{} creates a new channel to be sent as data, and sends it over channel $c$, and then terminates.
Each \sender{} process $\func{send} \, x$ and $\func{send} \, y$ is an instantiation of $\func{send} \, c$.
The \receiver{} receives data \emph{twice} on a channel $c$ and then terminates.
The \receiver{} process $\func{recv} \, z$ is an instantiation of $\func{recv} \, c$.
\todo{get rid of vertical space}
\begin{minipage}[t]{0.5\textwidth}
\begin{flalign*}
  & \func{send} \; : \; \Var_n \to \Process_n & \\
  & \func{send} \; c = \new{} \send{(\suc c)}{\constr{0}} \PO
\end{flalign*}
\end{minipage}
\begin{minipage}[t]{0.5\textwidth}
\begin{flalign*}
  & \func{recv} \; : \; \Var_n \to \Process_n & \\
  & \func{recv} \; c \; = \recv{c}{} \recv{(\suc c)} \PO
\end{flalign*}
\end{minipage}

The \courier{} role is defined below as $\func{carry} \; x \; y \; z$.
It sequentially receives on the two input channels $x$ and $y$ -- instantiating $in0$ and $in1$, and then outputs the two pieces of received data on the output channel $z$ -- instantiating $out$.
Finally, we create three communication channels and compose all four processes together: the first channel is shared between the one \sender{} and the \courier{}, the second between the other \sender{} and the \courier{}, and the third between the \receiver{}  and the \courier{}.
The result is the courier $\func{system}$ defined below.
\begin{minipage}[t]{0.5\textwidth}
\begin{flalign*}
  \func{carry} \; : \; \Var_n \to \Var_n \to \Var_n \to \Process_n%
\end{flalign*}%
\begin{flalign*}
  & \func{carry} \; in0 \; in1 \; out \; &&= \\
  & && \recv{in0 &&}{} \\
  & && \recv{(\suc \; in1) &&}{} \\
  & && \send{(\suc \suc \; out) &&}{\suc \constr{0}} \\
  & && \send{(\suc \suc \; out) &&}{\constr{0}} \PO
\end{flalign*}
\end{minipage}
\begin{minipage}[t]{0.5\textwidth}
\begin{flalign*}
  & \func{system} && : \; \Process_{\constr{0}} & \\
  & \func{system} && = \new{} ( \func{send} \; \constr{0} \\
  & && \comp{}{ \new{} ( \func{send} \; \constr{0} } \\
  & && \comp{}{ \new{} ( \func{recv} \; \constr{0} } \\
  & && \comp{}{ \func{carry} \; (\suc \suc \constr{0}) \; (\suc \constr{0}) \; \constr{0}))) } \\
\end{flalign*}
\end{minipage}
\todo{fix}

We continue this example and by providing typing derivations for these processes in \autoref{leftover-typing}.
There we use a mix of linear, graded and shared typing to type check $\func{system}$.
\end{example}
